---
title: "Biol 3872 Lab3-Assignment"
output: html_document
date: "`r Sys.Date()`"
author: "May Engelhardt"
student_id: "B00970301"
editor_options: 
  chunk_output_type: inline
---

**BEFORE STARTING:**\
1. Include your name (as shown in Brightspace) in the `author` section above as text (e.g., "Liam MacNeil").\
2. Install and load the packages required to run your code chunks/

Assignments will largely be evaluated based on your code chunk answers. If they are incomplete or formatted incorrectly (e.g., code chunks don't run), no marks will be awarded. I recommend experimenting with solving these answers in a separate script/markdown document and reporting your answers with minimal script.

```{r}
library(tidyverse)
library(ggplot2)
data("iris")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}\
\## 1. Task 1

A)  What species displays the longest petal lengths? Store the species name as a character called `longest`.
B)  What species has the greatest variance in sepal width? Store the species name as a character called `variable`.

Report your answer in the code chunks below.

```{r Task 1}
#A) 

max_val = max(iris$Petal.Length)
longest_row = iris[iris$Petal.Length == max_val, ]
longest = as.character(longest_row$Species)

#B)


Set = iris[iris$Species == "setosa",]
Vir = iris[iris$Species == "virginica",]
Ver = iris[iris$Species == "versicolor",]

var_set = var(Set$Sepal.Width)
var_vir = var(Vir$Sepal.Width)
var_ver = var(Ver$Sepal.Width)

results <- c("setosa" = var_set, "virginica" = var_vir, "versicolor" = var_ver)

variable <- names(results)[which.max(results)]

```

:::\

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 2. Task 2

A)  Using `base` R, produce a `plot` of all variables excluding the `species` variable but color points by species identity. Save this plot as a .png called `base_plot.png` in your `Figures` folder. Specify figure dimensions as 12 x 12 cm with 300 DPI.

```{r Task 2}
#A)
# YOUR CODE HERE



png("./Figures/base_plot.png",width=12,height=12,units="cm",res=300)
plot(iris[,1:4],col=iris$Species )
dev.off()

```
:::

\

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 3. Task 3

A)  Based on the grammar of graphics, produce a **blank plot (no data plotted)** named `blank` using the `iris` data set with species as the x axis and sepal widths as the y axis. Store this object as a `ggplot` object named `empty_plot`.

```{r Task 3}
#A)
empty_plot <- ggplot(
                    data=iris,
                    aes(x = Species, y = Sepal.Width)
                    )
empty_plot
```
:::

\

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 4. Task 4

A)  Fix the code below so that it runs and plots successfully and store it as a plot object named `fixed_traits_plot`. This plot should have petal length as the x-axis, sepal length as y axis and point size should correspond to petal width. Each point should be colored by species name.

```{r Task 4}
# FIX AND IMPLEMENT CODE BELOW
#ggplot(iris, aes(x = Petal.Length, y = Sepal.Lenth), color = "Species")+
 # geom_line(size = Petal.Width)+
  #theme_minimal()

# A)
fixed_traits_plot = ggplot(iris, aes(x = Petal.Length, 
                                     y = Sepal.Length, 
                                     col = Species, 
                                     size = Petal.Width
                                     )
                           )+
  geom_line()+
  theme_minimal()
```
:::

\

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 5. Task 5

A)  Fix and modify the code below and add a discretized `geom` (e.g., `hex`) with exactly 10 bins, with a `viridis` filled palette (check `?` for different `scale_fill_...` options), and facets for each species. Name and store it as a plot object as `fixed_traits_plot_discrete`. *Clue: Can you fill a hexagon with species names using a continuous palette?*

```{r Task 5}
# FIX AND IMPLEMENT CODE BELOW
#A) 

#ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))+
#  geom_line(alpha = 0.5, aes(as.factor(Petal.Length)))+
#  theme(legend.position = "left")+
#  scale_x_continuous(limits = seq(0,Inf,2))

fixed_traits_plot_discrete <- ggplot(iris, 
                                    aes(x = Sepal.Length, 
                                        y = Sepal.Width))+
  geom_hex(bins=10)+
  facet_wrap(~Species)+
  scale_fill_viridis_c()+
  scale_x_continuous(breaks = seq(0,10,2))+
    theme_minimal()+
  theme(legend.position = "left")



fixed_traits_plot_discrete  


```
:::

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 6. Task 6

A)  Using the hotkey to generate a pipe operator in the `tidyverse`, pivot the petal measurements in the `iris` dataset to a longer format and successfully generate a boxplot with these petal measurements as the y-axis and measurement type (e.g., length v. width) as the x axis.Use these unique color codes (`"forestgreen"`,`"skyblue"`,`"maroon"`) to **fill**- not color- each species. This should be generated in a single code chunk. Add the `minimal` theme with text size at 12.

Name and store it as a plot object as `petal_plot`.

```{r Task 6}
# YOUR CODE HERE

#A)
# YOUR CODE HERE
petal_plot <- iris %>% 
  pivot_longer(cols = c(Petal.Length,Petal.Width),
               names_to = "Measurement_Type",
               values_to = "Petal_Measurement") %>% 
  ggplot(aes(x=Measurement_Type, 
             y=Petal_Measurement, 
             fill=Species)) +
           geom_boxplot() +
           scale_fill_manual(values=c("forestgreen","skyblue","maroon")) +
           theme_minimal(base_size = 12)
         
         
petal_plot
```
:::

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 7. Task 7

A)  Similar to `Task 6`, in a single code chunk, pivot the petal measurements in the `iris` dataset to a longer format.

Generate a scatter plot comparing with different measurement types on the x axis and their values on the y axis.

Add a jitter `geom` and give each species a unique `facet` with a free (unique) y-axis.

Name and store it as a plot object as `facet_petal_plot`.

B)  Reproduce this same plot named `facet_petal_plot_reorder` where the facets are reversed by the reordering the `species` factor.

```{r Task 7}

#A)
# YOUR CODE HERE
facet_petal_plot <- iris %>% 
  pivot_longer(cols = c(Petal.Length,Petal.Width), names_to = "Measurement_Type",values_to = "Value") %>% 
  ggplot(aes(x=Measurement_Type, y=Value,col=Species))+
        geom_point()+
        geom_jitter()+
        facet_wrap(~Species, scales = "free_y")
facet_petal_plot

#B)
# YOUR CODE HERE
facet_petal_plot_reorder <- iris %>% 
   pivot_longer(cols = c(Petal.Length,Petal.Width), names_to = "Measurement_Type",values_to = "Value") %>% 
  mutate(Species = factor(Species, levels = rev(levels(iris$Species)))) %>% 
  ggplot(aes(x=Measurement_Type,y=Value,col=Species))+
  geom_point()+
  geom_jitter()+
  facet_wrap(~Species, scales = "free_y")

facet_petal_plot_reorder
```
:::

::: {style="background-color: #44AA99; transparent; padding: 5px; border-radius: 2px;"}
## 8. Task 8

A)  Create a new column in the `iris` dataset with `base` R or `dplyr` called `petal_area` which computes petal area.

B)  Create and store a ggplot object named `area_plot` comparing `petal_area` (x axis) to `Sepal.Length` (y axis) with points colored by species.

```{r Task 8}
#A)
iris_mod <- iris %>% 
  mutate(petal_area = Petal.Length*Petal.Width)

area_plot <- ggplot(data = iris_mod,aes(x=petal_area,y=Sepal.Length, col=Species))+
  geom_point()+
  theme_minimal()
area_plot
# YOUR CODE HERE
```
:::

## 8. Task 8C

C)  Save the previous plot as a `ggplot` object in your Figures folder named `area_plot.png` and define dimensions as 12x12 cm with 300 DPI.

```{r}
#C)
# YOUR CODE HERE


png("./Figures/area_plot.png",width=12,height=12,units="cm",res=300)
area_plot
dev.off()
```

Be sure before submitting this assignment to clear your and environment and re-run your code to ensure it works then submit this `markdown` file to the relevant Brightspace submission box.
