---
title: "Biol 3872 Lab4-Assignment"
author: "May Engelhardt"
student_id: "B00970301"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: inline
---

\
**BEFORE STARTING:**\
1. Include your name (as shown in Brightspace) in the `author` section above as text (e.g., "Liam MacNeil") and Banner ID ("B00123456") in `student_id`.\
2. Install and load the packages required to run your code chunks/

Assignments will largely be evaluated based on your code chunk answers. If they are incomplete or formatted incorrectly (e.g., code chunks don't run), no marks will be awarded. I recommend experimenting with solving these answers in a separate script/markdown document and reporting your answers with minimal script.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the packages needed to answer the questions below. If you realize you need a package, load (install if needed) here.
library(tidyverse)
library(RColorBrewer)
library(colorspace)
library(patchwork)
```

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 1 {2 points}

A)  Using the iris dataset, create a `ggplot` showing only the mean sepal length for each species using the `stat_summary` function and `geom` crossbar. Store in variable `iris_mean.`

B)  Now add a `geom` jitter layer to your plot showing how the data is distributed around the mean. Store in variable `iris_data.`

```{r}
data("iris")
#A)
iris_mean <- ggplot(iris, aes(x = Species, y = Sepal.Length)) +
  stat_summary(fun = "mean", geom = "crossbar")
iris_mean
#B)
iris_data <- iris_mean +
  geom_jitter() 
iris_data
# YOUR CODE HERE
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 2 {3 points}

Use the `mtcars` data set to compare fuel efficiency of car models and store as unique `ggplot` objects for each question named `fuel_efficiency_plot_a`, `fuel_efficiency_plot_b`, etc. Think about Tufte's rules 5, 7, 9.

A)  Transform rownames to a column titled `cartype`, convert this column to a factor ordered (descending) by the average `mpg` (miles per gallon), pipe this modified dataframe into a `ggplot` barplot.

B)  Add to `fuel_efficiency_plot_a`, a minimal theme with 12 font size, adjust the x labels to 45 degrees and horizontally (right) justify labels.

C)  Add to `fuel_efficiency_plot_b`, another column `geom` which is filled by car weight with a `viridis` color scheme.

```{r}
data("mtcars")
# A)
fuel_efficiency_plot_a <- mtcars %>% 
  rownames_to_column(var = "cartype") %>% 
    mutate(cartype = fct_reorder(cartype,mpg,.desc = T)) %>% 
  ggplot(aes(x = cartype, y = mpg)) +
  geom_col() +
  theme_dark()
fuel_efficiency_plot_a
         
# B)
fuel_efficiency_plot_b <- fuel_efficiency_plot_a +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1))
fuel_efficiency_plot_b
# C)
fuel_efficiency_plot_c <- fuel_efficiency_plot_b +
  geom_col(aes(fill=wt,)) +
  scale_fill_viridis_c()
fuel_efficiency_plot_c
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 3 {1 point}

A)  Use `patchwork` to combine a new plot with `fuel_efficiency_plot_c` called `horsepower_plot`. Execute similar data manipulation steps as in **Task 2** to convert rownames to an ordered column as a factor. Pipe this step into a ggplot object with horsepower as the y axis (Think: What would the x axis need to be to combine with `fuel_efficiency_plot_c`?), add a scatterplot `geom` with enlarged points (`size = 3`), a minimal theme with 12 point font, a modified x axis and color palettes identical to `fuel_efficiency_plot_c`.

Patch this `cars_compared` plot with `fuel_efficiency_plot_c` as a single column with two rows and combine the two x axes to avoid redundancy.

```{r}
# A)
horsepower_plot <- mtcars %>%
  rownames_to_column(var = "car_model") %>%
  mutate(car_model = factor(car_model, levels = car_model[order(mpg)])) %>% # Ordering factor by mpg to match Task 2
  ggplot(aes(x = car_model, y = hp, color = car_model)) +
  geom_point(size = 3) +
  theme_minimal(base_size = 12) +
  scale_color_discrete(guide = "none") + 
  labs(x = NULL, y = "Horsepower") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
cars_compared <- (fuel_efficiency_plot_c / horsepower_plot) + 
  plot_layout(guides = "collect", axes = "collect_x")

cars_compared
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 4 {2 points}

A)  Load in the CO2 data set used during the lab exercise and render it as a data frame called `co2_df` with columns `co2`, `year`, and `month`. Recall the steps taken during the exercise to properly format the data frame. Create a new column called `season` by three month intervals (Dec, Jan, Feb = Winter / Mar, Apr, May = Spring, etc.). Check out the `ifelse` statements from lab 2.

B)  Order the season variable as a factor (Levels = Winter, Spring, Summer, Autumn). Create a `ggplot` object named `seasonal_co2` plotting years on the x-axis and co2 concentration on the y-axis. Add a point `geom` colored by season with these colors ("#0072B2", "#009E73", "#CC79A7", "#E69F00"), a jitter `geom` to separate points for visualization, and a minimal `theme` with 12 point font.

```{r}
# A)
data("co2")
co2_df <- data.frame(
  co2 = as.vector(co2),
  year = as.integer(time(co2)),
  month = as.integer(cycle(co2))
)

# B)
co2_df$season <- ifelse(co2_df$month %in% c(12, 1, 2), "Winter",
                 ifelse(co2_df$month %in% c(3, 4, 5), "Spring",
                 ifelse(co2_df$month %in% c(6, 7, 8), "Summer",
                        "Autumn")))
co2_df$season <- factor(co2_df$season, 
                        levels = c("Winter", "Spring", "Summer", "Autumn"))

seasonal_co2 <- ggplot(co2_df, aes(x = year, y = co2, color = season)) +
  geom_point() +
  geom_jitter(width = 0.2, alpha = 0.6) + 
  scale_color_manual(values = c("#0072B2", "#009E73", "#CC79A7", "#E69F00")) +
  theme_minimal(base_size = 12) +
  labs(title = "Mauna Loa CO2 Concentration by Season",
       x = "Year", 
       y = "CO2 Concentration (ppm)")

seasonal_co2
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 5 {1 point}

A)  Using the same `co2_df` as above, order the months by the calender order (not alphabetical) as shown in the exercise. Create a `ggplot` object named `ordered_co2_friendly` comparing yearly co2 concentrations (same as Task 4). Add a point `geom` and a minimal `theme` with 12 point font and a manual color-blind friendly palette `scale` using `diverge_hcl` from `colorspace`. Finally, `facet` this plot by season.

```{r}
# A)
co2_df$month_name <- factor(month.abb[co2_df$month], levels = month.abb)

ordered_co2_friendly <- ggplot(co2_df, aes(x = year, y = co2, color = month_name)) +
  geom_point() +
  scale_color_manual(values = diverge_hcl(12)) +
  theme_minimal(base_size = 12) +
  facet_wrap(~season) +
  labs(title = "Yearly CO2 Concentrations by Season",
       x = "Year",
       y = "CO2 Concentration (ppm)",
       color = "Month")

ordered_co2_friendly
```
:::

\

**BEFORE SUBMITTING**

Clear your environment (broom) and run your entire script to ensure everything has been defined properly!

Be sure to submit this `markdown` file to the relevant Brightspace submission box.
