---
title: "Biol 3872 Lab6-Assignment"
output: html_document
date: "`r Sys.Date()`"
author: "May Engelhardt"
student_id: "B00970301"
---

\
**BEFORE STARTING:**\
1. Include your name (as shown in Brightspace) in the `author` section above as text (e.g., "Liam MacNeil") and Banner ID ("B00123456") in `student_id`.\
2. Install and load the packages required to run your code chunks/

```{r}
packages <- c("tidyverse", "lubridate", "nycflights13", "oce", "patchwork", "ggplot2")


# Install packages not yet installed
installed_packages <- packages %in% installed.packages()

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading

invisible(lapply(packages, library, character.only = TRUE))
```

Assignments will largely be evaluated based on your code chunk answers. If they are incomplete or formatted incorrectly (e.g., code chunks don't run), no marks will be awarded. I recommend experimenting with solving these answers in a separate script/markdown document and reporting your answers with minimal script.

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 1 {2 points}

A)  Convert and if needed, correct the character strings to dates 1-5 in unambiguous formats. Store these as variables with the given names.

B)  Convert and if needed, correct the character strings to datetimes 1-3 in unambiguous formats. Store these as variables with the given names. Give all datetimes in Atlantic Standard Time (AST). Check `?OlsonNames`.

```{r}
# YOUR CODE HERE

# A)
# Roald Amundsen reaches the South Pole
date1 <- as.Date("Dec/14/1911", format = "%b/%d/%Y")

# Alexander Graham Bell and Thomas Watson exchange the first telephone call
date2 <- as.Date("January-25-1915", format = "%B-%d-%Y")


# Sputnik is launched into orbit
date3 <- as.Date("Oct-4-1957", format = "%b-%d-%Y")

# The first Woodstock festival begins
date4 <- as.Date("August-16-69", format = "%B-%d-%y")

# The first Blackberry mobile device is released by Research In Motion
date5 <- as.Date("19/99/January", format = "%d/%y/%B")

# B)


# The estimated minute (Pacific Standard Time) Sidney Crosby scored the overtime goal to win gold in 2010 Olympics
datetime1 <- as.POSIXct("February-28-2010 02:38PM", format = "%B-%d-%Y %I:%M%p", tz = "Canada/Atlantic")

# The timestamp of the first observations detecting gravitational waves. These were produced by the whirling merger of two black holes, generating ripples in space-time confirming Einstein's prediction from general relativity (reported in UTC; observed first at 05:50:45 in Louisiana within Eastern Standard Time)
datetime2 <- as.POSIXct("Sep/14/2015 09:50:45", format = "%b/%d/%Y %H:%M:%S", tz = "Canada/Atlantic")


# The beginning minute a total eclipse* was visible over large sections of Nova Scotia
# * These occur every ~18 months, but in any one (inhabited) place, may not occur for centuries.
datetime3 <- as.POSIXct("04-8-2024 16:39", format = "%m-%d-%Y %H:%M", tz = "Canada/Atlantic")

```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 2 {3 points}

A)  Using the Bedford Basin data (`Bedford_ts`) and your skills indexing data, extract all the rows where `year` equals 2016, store this data in a variable called `ts_2016`. **Do you need to load certain packages to complete this?**

B)  Convert the `month` column to full month names in column called `month_full`.

C)  Summarize microbial richness (median) by each year and each month, then plot it using `ggplot` with the x-axis showing full month names (**arranged by calendar order**), coloring points by `Year`. Store this plot in a plot object called `monthly_richness`.

**NOT MARKED**: Make this plot pretty: e.g., modify x-axis labels to be readable, edit `Year` to be discrete, apply a nice color palette, and customize in other ways that improve plot quality.

```{r}
library(tidyverse)

Bedford_ts <- read.csv("../Data/BedfordBasin_ts.csv") %>%
  mutate(Year = year(Date),
         Month = month(Date))

# YOUR CODE HERE
# A)
ts_2016 <- Bedford_ts[Bedford_ts$year == 2016]


# B)
Bedford_ts$month_full <- month.name[Bedford_ts$Month]


# C)

summary_df <- Bedford_ts %>%
  mutate(
    Year = factor(year(Date)),
    month_full = month(Month, label = TRUE, abbr = FALSE)
  ) %>%
  group_by(Year, month_full) %>%
  summarize(Median_Richness = median(Richness, na.rm = TRUE), .groups = "drop")

monthly_richness <- ggplot(summary_df, aes(x = month_full, y = Median_Richness, color = Year)) +
  geom_point(size = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

monthly_richness
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 3 {3 poinxts}

A)  Take the sequences of `day`, `month`, `year`, `hour`, `min`, and `secs` and combine them into a single vector of unambiguous datetimes named `combined_datetime`.

B)  Search `OlsonNames` to find the time zone for Stockholm (Sweden) and assign this to datetimes in a new object called `combined_datetime_stock`.

C)  Convert these times to the Atlantic Canada timezone (AST) and store named these datetimes in `combined_datetime_NS`.

```{r}
day <- seq(1,30, 1)
month <- rep(seq(1, 10, 1), 3)
year <- 1978:2007
hour <- rep(11:20, 3)
min <- 15:44
sec <- 4:33

# A)
combined_datetime <- make_datetime(year, month, day, hour, min, sec)

# B)
combined_datetime_stock <- force_tz(combined_datetime, tzone = "Europe/Stockholm")
# C)
combined_datetime_NS <- with_tz(combined_datetime_stock, tzone = "America/Halifax")
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 4 {3 points}

A)  Convert vector of `dates` to a date object called `dates_format`.

B)  Calculate the difference between each date in sequence (date 1 to date 2, date 2 to date 3, etc.) and store this as a **vector** called `dfftime_v1`. This can be done manually, you can also check functions from `lubridate` (`?diff`), but remember to convert to a vector.

C)  Index `dates_format` to remove all dates which are after the date when gravitational waves were first measured (see Q1). Store this as a vector called `pre_Einstein`.

```{r}
dates <- c("1974-01-01", "1981-09-04", "1993-05-18", "2004-12-11", "2015-09-14", "2023-01-13")

# YOUR CODE HERE
# A)
dates_format <- as.Date(dates)
# B)
dfftime_v1 <- as.vector(diff(dates_format))
# C)
gw_date <- as.Date("2015-09-14")
pre_Einstein <- dates_format[dates_format <= gw_date]
```
:::

\

::: {style="background-color: #44AA99; padding: padding: 5px; border-radius: 2px;"}
## Task 5 {2 points}

A)  Using the Bedford Basin data (`Bedford_ts`), create a `for()` loop or use an `*apply` that calculates the weekly median microbial richness and store it in a variable called `weekly_median`.

B)  Create a `ggplot` object called `Richness_weekly` which plots raw richness (y-axis) against weeks (x-axis). Color points based on their `month_full` name (ordered by calender). Overlay `weekly_median` into this plot with a unique `geom_point` layer.\

```{r}
# YOUR CODE HERE
# A)
rich_vec <- as.numeric(unlist(Bedford_ts$Richness))
date_vec <- as.Date(unlist(Bedford_ts$Date))
week_vec <- week(date_vec)
month_vec <- factor(unlist(Bedford_ts$month_full), levels = month.name)

weekly_median_vals <- tapply(rich_vec, week_vec, median, na.rm = TRUE)
weekly_median <- data.frame(
  week = as.numeric(names(weekly_median_vals)),
  Richness = as.vector(weekly_median_vals)
)

Bedford_df <- data.frame(
  week = week_vec,
  Richness = rich_vec,
  month_full = month_vec
)
```

```{r}

# B)
Richness_weekly <- ggplot(Bedford_df, aes(x = week, y = Richness)) +
  geom_point(aes(color = month_full), alpha = 0.6) +
  geom_point(data = weekly_median, col = "black", size = 3) +
  labs(title = "Microbial Richness by Week",
       x = "Week of Year",
       y = "Richness",
       color = "Month") +
  theme_minimal()

Richness_weekly
```
:::

Clear your environment (broom) and run your entire script to ensure everything has been defined properly!

Be sure to submit this `markdown` file to the relevant Brightspace submission box.
